### 2018-04-30 4_impl.tex
add some text on Y-tree.

Therefore, while working on the new design of \porthos[2], we decided to clearly separate the high-level intermediate code representation (implemented as a recursive AST structure) from low-level event-based representation (implemented as an event-flow graph).
Such a modular architecture allows to support multiple input languages by parsing them and converting parsed syntax trees to a simplified AST. which, along with all other data-transfer objects (DTO), must be immutable, so that it is possible to guarantee the correctness of the program by controlling its invariants.
The immutability in \porthos[2] is implemented via \texttt{final} fields that are assigned by the immutable-object values (either a primitive type, or another immutable object, or an immutable collection provided by the library Guava by Google%
\footnote{Project repository: \url{https://github.com/google/guava/}}%
).

@@

As \porthos[1], the \porthos[2] uses the open-source SMT solver \texttt{Z3} from Microsoft Research~\ref{de2008z3}. However, unlikely its predecessor, the \porthos[2] has an additional abstraction level \texttt{Z-formula} (see Section~\ref{ch:impl:model:zformula}) that allow to use any other SMT solver. %TODO: *should* allow?

@@

The general architecture scheme of \porthos[2] is presented in Figure~\ref{fig:arch}.
The program takes as input the program to be analysed and one (the reachability analysis mode) or two (the portability analysis mode) memory models. The parsed program syntax tree is then converted (Section~\ref{ch:impl:proc:y-constr}) to a program AST called \texttt{Y-tree}%
\footnote{In order to avoid confusion between different internal representations, we prefix the names of elements of each internal representation with a letter. For instance, we picked the letter `Y' to denote the AST code representation as drawing of this letter resembles the tree branching; with letter `X' we prefix elements of the event-flow graph as the events are to be e\textbf{x}ecuted; and with letter `W' we prefix elements of the \textbf{w}eak memory model AST.} %
(Section~\ref{ch:impl:model:ytree})
%, and the memory model is being parsed (Section~\ref{ch:impl:proc:w-parser}) to another AST called \texttt{W-model} (Section~\ref{ch:impl:model:wmodel}).
, which then is being pre-processed at the pre-compilation stage (Section~\ref{ch:impl:proc:x-pre-compiler}) in order to collect information necessary for the compilation.
The \texttt{Y-tree} then is being compiled (Section~\ref{ch:impl:proc:x-compiler}) to an \texttt{X-graph} representation (Section~\ref{ch:impl:model:xgraph}).
The compiled \texttt{X-graph} then undergoes a number of transformations (Section~\ref{ch:impl:proc:x-transf}) necessary for the encoding into a \texttt{Z-formula} (Section~\ref{ch:impl:model:zformula}) at the pre-encoding stage (Section~\ref{ch:impl:proc:x-transf}).
Then, the memory-model constructor (Section~\ref{ch:impl:proc:w-constr}) constructs the derived relations on the basis of the \texttt{X-graph} in order to build the weak memory model \texttt{W-model} (Section~\ref{ch:impl:model:wmodel}).
Thereafter, \texttt{W-model} and transformed \texttt{X-graph} are encoded (Section~\ref{ch:impl:proc:z-encoder}) to a \texttt{Z-formula} representation (a wrapper over an SMT logical formula), which then is passed as an input to the SMT-solver.

@@

Figure~\ref{fig:in_grammar_pts} represents the simplified grammar of of input language used by previous version of \porthos{} (the full ANTLR grammar is available at Appendix~\ref{apx:in_grammar_pts}.

\begin{figure}[H]
\begin{lstlisting}[mathescape=true,%
                  caption={The sketch of the input language grammar used by \oldporthos},%
                  label={fig:in_grammar_pts},%
                  morekeywords={if,then,else,return,while,program,thread},%
                  breaklines=true,%
                  basicstyle=\ttfamily\scriptsize]
<prog> : <init> <thrd>* <assert>
       ;
<thrd> : thread <tid> <inst>
       ;
<inst> : <atom>
       | <inst> ; <inst>
       | while <pred> <inst>
       | if <pred> { <inst> } <inst>
       ;
<atom> : <reg> <- <expr>
       | <reg> <:- <loc>
       | <loc> := <reg>
       | <reg> = <loc>.load(<atomic>)
       | <loc> = <reg>.store(<atomic>)
       | 'mfence'
       | 'sync'
       | 'lwsync'
       | 'isync'
       ;
<pred> :
       | true
       | false
       | <expr> (and | or) <expr>
       | <expr> ('==' | '!=' | '>' | '>=' | '<' | '>=') <expr>
       ;
<expr> : [0-9]
       | <reg>
       | <expr> ('*' | '+' | '-' | '/' | '%') <expr>
       ;
\end{lstlisting}
\end{figure}

@@

The minor drawbacks of grammar used by \porthos[1] include lack of operator associativity (expressions of type \lstinline{1 + 2 * 3} could not be parsed), incorrectly (in terms of C) implemented grammar rule for statements (the semicolon punctuator `\lstinline{;}' was implemented as the separator between two statements, whereas in C it serves as a statement terminator). Also, \porthos[1] supports only the litmus-specific syntax for the variables initialisation, however, it allows only initialisation of the shared variable and only by default value \lstinline{0}. The \porthos[2] supports arbitrary declaration to be performed in initialisation statement.

@@

\subsection{The internal representations}
\label{ch:impl:model}

%As it has already been mentioned, all internal representations constructed by \porthos[2] are \textit{immutable}.

@@

@@

>>>>

Therefore, while working on the new design of \porthos[2], we decided to clearly separate the high-level intermediate code representation (implemented as a recursive AST structure) from low-level event-based representation (implemented as an event-flow graph).
Such a modular architecture allows to support multiple input languages by parsing them and converting parsed syntax trees to a simplified AST. which, along with all other data-transfer objects (DTO), must be immutable, so that it is possible to guarantee the correctness of the program by controlling its invariants.
The immutability in \porthos[2] is implemented via \texttt{final} fields that are assigned by the immutable-object values (either a primitive type, or another immutable object, or an immutable collection provided by the library Guava by Google%
\footnote{Guava project repository: \url{https://github.com/google/guava/}}%
).

@@

As \porthos[1], the \porthos[2] uses the open-source SMT solver \texttt{Z3} from Microsoft Research~\cite{de2008z3}. However, unlikely its predecessor, the \porthos[2] has an additional abstraction level \zformula{} (see Section~\ref{ch:impl:model:zformula}) that allow to use any other SMT solver. %TODO: *should* allow?

@@

The general architecture scheme of \porthos[2] is presented in Figure~\ref{fig:arch}.
The program takes as input the program to be analysed and one (the reachability analysis mode) or two (the portability analysis mode) memory models. The parsed program syntax tree is then converted (Section~\ref{ch:impl:proc:y-constr}) to a program AST called \ytree{}%
\footnote{In order to avoid confusion between different internal representations, we prefix the names of elements of each internal representation with a letter. For instance, we picked the letter `Y' to denote the AST code representation as drawing of this letter resembles the tree branching; with letter `X' we prefix elements of the event-flow graph as the events are to be e\textbf{x}ecuted; and with letter `W' we prefix elements of the \textbf{w}eak memory model AST.} %
(Section~\ref{ch:impl:model:ytree})
%, and the memory model is being parsed (Section~\ref{ch:impl:proc:w-parser}) to another AST called \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
, which then is being pre-processed at the pre-compilation stage (Section~\ref{ch:impl:proc:x-pre-compiler}) in order to collect information necessary for the compilation.
The \ytree{} then is being compiled (Section~\ref{ch:impl:proc:x-compiler}) to an \xgraph{} representation (Section~\ref{ch:impl:model:xgraph}).
The compiled \xgraph{} then undergoes a number of transformations (Section~\ref{ch:impl:proc:x-transf}) necessary for the encoding into a \zformula{} (Section~\ref{ch:impl:model:zformula}) at the pre-encoding stage (Section~\ref{ch:impl:proc:x-transf}).
Then, the memory-model constructor (Section~\ref{ch:impl:proc:w-constr}) constructs the derived relations on the basis of the \xgraph{} in order to build the weak memory model \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
Thereafter, \wmodel{} and transformed \xgraph{} are encoded (Section~\ref{ch:impl:proc:z-encoder}) to a \zformula{} representation (a wrapper over an SMT logical formula), which then is passed as an input to the SMT-solver.

@@

Figure~\ref{fig:in_grammar_pts} represents the simplified grammar of of input language used by previous version of \porthos{} (the full ANTLR grammar is available at Appendix~\ref{apx:grammar}.

\begin{figure}[h]%[!hb]
\centering
\begin{lstlisting}[mathescape=true,%
                  caption={The sketch of the input language grammar used by \oldporthos},%
                  label={fig:in_grammar_pts},%
                  morekeywords={if,then,else,return,while,program,thread},%
                  breaklines=true,%
                  basicstyle=\ttfamily\scriptsize]
<prog> : <init> <thrd>* <assert>
       ;
<thrd> : thread <tid> <inst>
       ;
<inst> : <atom>
       | <inst> ; <inst>
       | while <pred> <inst>
       | if <pred> { <inst> } <inst>
       ;
<atom> : <reg> <- <expr>
       | <reg> <:- <loc>
       | <loc> := <reg>
       | <reg> = <loc>.load(<atomic>)
       | <loc> = <reg>.store(<atomic>)
       | 'mfence'
       | 'sync'
       | 'lwsync'
       | 'isync'
       ;
<pred> :
       | true
       | false
       | <expr> (and | or) <expr>
       | <expr> ('==' | '!=' | '>' | '>=' | '<' | '>=') <expr>
       ;
<expr> : [0-9]
       | <reg>
       | <expr> ('*' | '+' | '-' | '/' | '%') <expr>
       ;
\end{lstlisting}
\end{figure}

@@

The minor drawbacks of grammar used by \porthos[1] include lack of operator associativity (expressions of type \lstinline{1 + 2 * 3} could not be parsed), incorrectly (in terms of C) implemented grammar rule for statements (the semicolon punctuator `\lstinline{;}' was implemented as the separator between two statements, whereas in C it serves as a statement terminator). Also, \porthos[1] supports only the litmus-specific syntax for the variables initialisation, however, it allows only ini
tialisation of the shared variable and only by default value \lstinline{0}. The \porthos[2] supports arbitrary declaration to be performed in initialisation statement.

@@

\subsection{The internal representations}
\label{ch:impl:model}
%As it has already been mentioned, all internal representations constructed by \porthos[2] are \textit{immutable}.

@@

- TODO: everything inherited YEntity + citation

The first internal representation that \porthos[2] uses is the \ytree{}, which represents a rather high-level AST.
The Apendix~\ref{apx:trees} represents the file tree of main classes that constitute the \ytree{} hierarchy (as the inheritance tree might be obvious for the C-like AST, we confine ourselves to representing the classes file tree only).

%\columnratio{0.7}%\begin{paracol}{2}
%\switchcolumn%\VerbatimInput{inc/lst/Ytree-tree.out}%\end{paracol}
The abstract syntax tree \ytree{} is an abstraction level suitable for compiling to a low-level representation (in case of processing low-level assembly code, it may be directly converted to the \xgraph{} representation).
As the ANTLR syntax tree follows the same structure as the grammar, which is a superset of the real (meaningful) grammar of C language, it lacks multiple concepts of the language (for example, the syntax of indexer access corresponds the grammar rule \lstinline{postfixExpression '[' expression ']'}, that is converted to the \texttt{YIndexerExpression}).
However some details of the syntax might have been abstracted away (for instance, array operations may be emulated by functions invocations, see~\cite{gries2012science}\textbf{TODO:CHAPTER,PAGE}), we found this level of abstraction as suitable enough for our tasks.

Following the terms of C standard, we distinguish \textit{statements} (actions to be performed) and \textit{expressions} (TODO).

- TODO: all statements inherit \texttt{YStatement} + interface lstlisting

The \porthos[2] processes the following statements: 
\begin{itemize}[nolistsep]
  \item \texttt{YBranchingStatement} representing the \texttt{if-then-else} statement,
  \item \texttt{YLoopStatement} representing both \texttt{while}- and \texttt{for}-loops, %TODO:RENAME YWhileLoopStatement->YLoopStatement in code!!!)}
  \item \texttt{YJumpStatement} representing unconditional jump (\texttt{goto}-jump to a label and loop-jumps \texttt{break} and \texttt{continue}),
  \item \texttt{YCompoundStatement} (block statement) representing sequence of \texttt{N} statements grouped into one syntactic unit,
  \item \texttt{YLinearStatement} representing single expression.
\end{itemize}

According to the C standard, \textit{"any statement may be preceded by a prefix that declares an identifier as a label name"}.
However, the \ytree{} statements may have only simple symbolic labels, which have to be resolved at the pre-compilation stage.
Apart from the listed set of statements, \ytree{} has also the \texttt{YFunctionDefinition} and its litmus-specific inheritor \texttt{YProcessStatement} used in intra-procedural analysis mode.
The function definition contains the \texttt{YCompoundStatement}-typed body and the signature of type \texttt{YMethodSignature}, which is used in the function resolution during the compilation stage. %TODO: pre-compilation?
The other litmus-specific statements are \texttt{YPreludeStatement} that carries the list of \texttt{YStatement}-typed initial writes, and \texttt{YPostludeStatement} that carries the \texttt{YExpression}-typed binary expression to be asserted by the litmus test.

Similar to statements of \ytree{}, its expressions

pointer operations are modeled as the pointer level of and expression (although it is the property of type, the y-tree is untyped, so elements that represent syntactic expressions shoudl carry tnis information)

Each element of \ytree{} carries the \texttt{OriginLocation} %
\textbf{(TODO: Rename CodeLocation->OriginLocation in the code!)} %
instance that contains information about the coordinates of the input text that generated the \ytree{} element.

@@

The ANTLR syntax tree of C

%, comparing to ANTLR syntax tree, does not have 
- desugar, equiv transform
- variables distinguished from
